Cognitive processes are robust to early environmental conditions in two lizard species

### 2024-10-31 ms.qmd
Updates

## Abstract
Animals must acquire new information through learning to adjust their behavior adaptively. However, learning ability can be constrained by conditions experienced during the early stages of development, when the brain is especially susceptible to environmental inputs. Temperature can result in phenotypically plastic adjustments to growth, metabolism, and learning in ectotherms. In vertebrates, abrupt thermal changes can also trigger chronic stress, mediated by glucocorticoids (GCs) that can be transmitted to offspring, potentially impacting their learning abilities. GCs and thermal environments are, therefore, predicted to have interactive effects on the development of learning in ectotherms. Here, we investigated the combined effects of prenatal corticosterone (CORT) - the main GC in reptiles - and incubation temperature on associative learning using two species of lizards, _Lampropholis delicata_ and _L. guichenoti_. We manipulated CORT levels and temperature in a 2x2 factorial design, and then subjected lizards to a color-associative learning task. We predicted that elevated CORT and low temperatures would impair associative learning. However, both species showed similar learning rates independently of treatment. Our results suggest that these two species may have evolved mechanisms to maintain learning performance despite prenatal challenges. We also found that color affected decision-making in both species. Overall, we observed a preference towards blue, underscoring the need to carefully select the color used in cognitive tests involving visual stimuli.

## Introduction
Cognition is a set of processes by which animals gather, preserve, and use information from their environment through perception, learning, memory, and decision-making [@shettleworth]. These processes are fundamental to foraging, mate selection, antipredatory strategies, and social behaviors, all of which are crucial for survival and reproduction [@dukas_evolutionary_2004]. Learning - acquiring and consolidating new information [@dukas_evolutionary_2004] - is essential for coping with environmental changes by enabling individuals to create new associations between events [@dukas_evolutionary_2004; @leal_behavioural_2012; @buchanan_condition_2013]. However, the capacity to acquire information varies among individuals, shaped by factors such as age, sex, and the developmental environment [@szuran_water_1994; @lemaire_prenatal_2000; @zhu_prenatal_2004; @amiel_hotter_2012; @amiel_egg_2014; @carazo_sex_2014; @noble_age-dependent_2014]. Interindividual differences in learning can lead to unequal responses to environmental conditions, potentially affecting population dynamics [@ward2016ecological; @welklin2024spatial]. For example, fast learners may better exploit novel resources or avoid new dangerous stimuli, while those with lower learning capabilities might struggle to adapt to environmental changes. Learning rate could affect their survival and reproductive output, ultimately influencing population growth rates and stability [@ward2016ecological; @welklin2024spatial]. Therefore, understanding the factors that shape learning is crucial to predict how populations will respond to novel circumstances.

@@

Developmental conditions are especially relevant in shaping cognitive abilities because the brain is highly susceptible to environmental inputs during early stages of life [@zhu_prenatal_2004]. Impacts on brain development can have long-lasting effects on cognition, potentially influencing an individual's capacity to learn and adapt to new environments [@lemaire_prenatal_2000; @zhu_prenatal_2004; @amiel_hotter_2012; @abayarathna_effects_2020]. For example, the state of the mother can influence offspring phenotype beyond genetic transmission through nest-site selection, provisioning, or the transmission of hormones or nutrients [i.e., 'maternal effects' @moore_mothers_2019]; and the developing brain is also susceptible to these effects.

Glucocorticoids (GCs) are a class of steroid hormones that are particularly relevant in this context. In vertebrates, GCs regulate metabolism and maintain homeostasis in response to disturbances [i.e., ‘the stress response’ @sapolsky_how_2000]. Under stressful situations, animals react by initiating adaptive physiological and behavioral adjustments mediated by GCs. These GCs can be transmitted directly from parents to their offspring with various effects on offspring phenotype [@Crino_2023]. Elevation in GCs during early stages of development typically results in altered neurogenesis, brain structure, and metabolic activity that, in most cases, are related to decreased learning abilities [@lemaire_prenatal_2000; @zhu_prenatal_2004; @du_dynamic_2009; @eaton2015mild; @farrell_developmental_2015-learn; @farrell_developmental_2016]. For instance, prenatal stress in rats (_Rattus norvegicus_) suppresses neurogenesis in the dentate gyrus, which is associated with impairments in hippocampal-related spatial tasks [@lemaire_prenatal_2000]. Factors such as sex or the nature of the learning task can also affect the direction and magnitude of the effects of prenatal exposure to elevated GCs [@szuran_water_1994; @crino_corticosterone_2014-learn; @farrell_developmental_2015-learn; @bebus_associative_2016; @farrell_developmental_2016]. Because the effects of prenatal GCs on learning can be context-dependent, it is crucial to conduct studies across diverse taxa and experimental conditions to understand these effects fully.

In addition to the environments experienced by parents, offspring also experience potentially stressful environmental conditions that can interact with or amplify parental effects. In ectotherms, the early thermal environment is a pervasive mechanism of phenotypic plasticity, influencing a broad spectrum of traits, including growth, metabolism, or cognition [@amiel_hotter_2012; @amiel_egg_2014; @dayananda_incubation_2017; @noble_developmental_2018; @abayarathna_effects_2020]. For instance, Port Jackson sharks (_Heterodontus portusjacksoni_) incubated at 23.6ºC took fewer days to master a numerical learning task than those incubated at 20.6ºC [@vila_pouca_quantity_2019]. In skinks, high incubation temperatures are generally associated with faster learning rates [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014]; but velvet geckos (_Amalosia lesueurii_) incubated at temperatures beyond their natural thermal range are worse learners than those incubated at colder temperatures [@dayananda_incubation_2017; @abayarathna_effects_2020]. Changes in neural structure and function likely mediate the effects of incubation temperature in reptiles, as high temperatures increase neural density and metabolic activity in the brain [@coomber_independent_1997; @sakata_neural_2000; @amiel_effects_2017; @beltran_are_2021].

GCs can play a pivotal role in determining vertebrate responses to elevated temperatures [@Crino_2023], potentially fostering natural interactions between temperature and GCs. Additionally, GCs and temperatures can act upon similar physiological mechanisms [@coomber_independent_1997; @lemaire_prenatal_2000; @sakata_neural_2000; @zhu_prenatal_2004; @du_dynamic_2009; @amiel_effects_2017; @beltran_are_2021]. Therefore, the effects of GCs and temperature may be interdependent, with the impact of one factor being modulated by the other. However, the interactive effects of GCs and temperature on learning abilities remain unexplored, yet understanding of how temperature and GC’s interact is particularly relevant in global climate change [@vinogradov2024inbreeding].

Here, we investigated the combined effects of prenatal corticosterone (CORT) - the main GC in reptiles [@Crino_2023] - and incubation temperature on cognition in two species of skinks, the delicate skink (_Lampropholis delicata_) and the common garden skink (_L. guichenoti_). We manipulated CORT levels in the eggs and then incubated them at two different temperatures in a 2X2 factorial design. Post-incubation, juveniles were subjected to a color-associative task to assess their learning abilities. We hypothesized that prenatal CORT levels and thermal environment would induce sustained effects on brain physiology, ultimately impacting the learning of an association task. We predicted that individuals exposed to high levels of CORT or low temperatures would learn slower than control individuals or those incubated at higher temperatures [@zhu_prenatal_2004; @amiel_hotter_2012; @eaton2015mild; @amiel_effects_2017]. Additionally, we predicted that incubation at high temperatures would mitigate the impact of CORT on learning performance, while cold incubation temperatures were expected to enhance the effects of CORT. The interactive effects of CORT and temperature may occur for two reasons that are not mutually exclusive: first, an increase in temperature leads to an overall rise in neural density [@amiel_effects_2017], thereby counteracting the impact of CORT [see @lemaire_prenatal_2000; @zhu_prenatal_2004; @eaton2015mild]; and second, the elevated metabolic rate associated with higher temperatures can accelerate CORT metabolism, resulting in embryos being exposed to CORT for a shorter time.

@@

*Eggs collection and incubation* -- Between mid-October 2022 to the end of February 2023, we provided females with a place to lay eggs by means of small boxes (12.5 L x 8.3 W x 5 H cm) containing moist vermiculite. These boxes were placed in one side of the communal enclosures (see above). We checked for eggs in the boxes three days a week (Monday, Wednesday, and Friday). After collection, we measured the length and width of eggs with a digital caliper to the nearest 0.1 mm, and mass using a digital scale (OHAUS, Model spx123; ± 0.001 g error). Then, we assigned individual IDs to the clutch and each egg. Eggs were treated with CORT or vehicle (see CORT and Temperature Manipulation below) and were placed in individual cups (80 mL) with moist vermiculite (12 parts water to 4 parts vermiculite). The cups were covered with cling wrap to retain moisture and left in LATWIT 2X5D-R1160 incubators at two different temperatures (see CORT and Temperature manipulation below) until hatching. Eggs were then checked three times a week for hatchlings.

*Hatchlings* -- After hatching, we measured juveniles' snout-to-vent length (SVL) and tail length (TL) with a ruler to the nearest mm, and mass using a digital scale (OHAUS, Model spx123; ± 0.001 g error). Hatchlings were then housed individually in small enclosures (18.7L x 13.2W x 6.3H cm) provided with non-stick matting, shelter, and a water dish. All care otherwise follows similar protocols to adults (see above).

Two weeks prior to the training phase (see details below), lizards were moved to the experimental arena for acclimatization. The arenas were individual medium-size (41 L x 29.7 W x 22 H cm) plastic containers with a shelter (9 L x 6 W x 1.5 H cm) on one of the sides and a water dish on the other. These arenas were placed in two rooms in 7 different racks that were monitored by 7 different CCTV systems (device model DVR-HP210475), allowing us to record their behavior during the experiment (see details below). Although the conditions in the experimental rooms were identical to the colony room, the number of lizards per species and treatment in each rack was counterbalanced to control for any potential effect of the room or the position of the lizard in the rack. During acclimatization and throughout the experiment, lizards were fed with only one cricket per day dusted with calcium and multivitamins, and water was supplied *ad libitum*. We provided a temperature gradient by means of a heat cord and heat lamps in a 12 h light: 12 h dark cycle.

#### CORT and Temperature Manipulation  
To empirically test the effect of early environment, we manipulated CORT concentration in eggs and incubated them under one of two temperature regimes (‘Cold’ - 23ºC ± 3ºC or ‘Hot’ - 28ºC ± 3ºC) in a 2x2 factorial design (@fig-Methods A). We topically dosed eggs with a CORT solution (10pg/mL) or a control treatment (100% ethanol). Corticosterone treatments were made by dissolving crystalline corticosterone (Sigma, Cat. No. C2505) in 100% ethanol. To dose eggs, we applied 5µl of solutions to eggshells using a micropipette. We selected these doses based on previous studies publishing yolk CORT concentrations in other oviparous reptiles [@lovern2008effects; @hanover2019corticosterone], while also validating that it fell within the range of CORT concentrations in eggs within our population [@crino2024eggs]. CORT treatment increased mean yolk CORT levels ~3.7x higher than control eggs [@crino2024eggs]. After the treatment, the eggs were incubated in one of the two previously mentioned temperature regimes (‘Cold’ or ‘Hot’) until hatching. The number of eggs per clutch assigned to each hormone and temperature treatment was counterbalanced in both species.

#### Learning  
To estimate learning skills, we tested each lizard's ability to locate a food reward in a color-associative learning task (@fig-Methods C, D). First, we performed a training phase where lizards had to learn to eat from white 3D-printed PLA ramps (9 L x 4 W x 5 H cm) identical in size and shape to the ones used during the associative task. We divided the training phase into three stages: the first stage where lizards had to eat a small, frozen cricket (*A. domestica*) from an opaque petri dish (3D x 1.6H cm) placed in the middle of their enclosure (@fig-Methods C, Stage 1); the second stage where the petri dish with the cricket was placed on top of the white 3D printed ramps (@fig-Methods C, Stage 2); and the third where the cricket was left inside a well (3D x 1.75H cm) on the top of the ramp (@fig-Methods C, Stage 3). Every trial began when we left the feeding block (petri dish, ramp, or both) inside the enclosure and finished one hour later when we took it away. At the end of each trial, we recorded whether the cricket had been consumed. A trial was considered successful if the lizard could locate and consume the reward, while completion of each stage required the lizards to eat the crickets in at least 5 out of 6 trials to ensure lizards were feeding consistently. This phase lasted 38 days until all the lizards learned to eat from the ramp; only in one case did we decide not to use the lizard because its behavior inconsistent throughout  the course of the training phase.

@@

```{r, fig-Methods}
#| label: fig-Methods
#| fig.cap: "Experimental design of the experiment. Panel A shows the early environment manipulation (see Methods). In panel B, the measurements of the 3D-printed ramps that were employed in the habituation and learning tasks. Panel C shows the habituation process in three different stages. In panel D, the associative task is done with the three different 3D-printed ramps. White lids in D show the ramps where the food reward was not attainable."

@@

We started with 96 lizards, 48 per species and 12 per treatment per species. However, due to natural mortality (n = 11), failing to pass the training stage (n = 1), or lack of motivation during the learning tasks (n = 3), we had a final sample size of 81 lizards. Final sample sizes per treatment and species are listed in @fig-deli and @fig-guich (figures for both species with the raw data are included in the _Supplementary Material_). Mean slopes (denoted as β throughout) per treatment for both species are provided in Table 1 in the _Supplementary Material_.    
```{r, models}
#| label: models
# Fitting the model and extraction of posteriors for both species using fit_m function (see func.r in R folder). The result everytime the function is used is a df with the posteriors of the model. The functions saves the model automatically in output/models; and when the parameter refit = FALSE then the posteriors are extracted from the model previously written instead of fitting the model again each time.
source(here("R", "func.R"))
# Model formula: FC_associative ~ Trial*cort*temp + (1 + Trial|lizard_id) + (1| clutch)
## A.1) L. delicata
deli_red <- fit_m("deli", "red", refit = FALSE)
write.csv(deli_red, file= "./output/Checking/deli_red.csv")
deli_blue <- fit_m("deli", "blue", refit = FALSE)
write.csv(deli_blue, file= "./output/Checking/deli_blue.csv")
## A.2) L. guichenoti
guich_red <- fit_m("guich", "red", refit = FALSE)
write.csv(guich_red, file= "./output/Checking/guich_red.csv")
guich_blue <- fit_m("guich", "blue", refit = FALSE)
write.csv(guich_blue, file= "./output/Checking/guich_blue.csv")
```

@@

_Lampropholis delicata_: _L. delicata_ that were given the blue ramp as the correct choice had an `r format_dec(((mean(c(probright_dbCORTCold, probright_dbControlCold, probright_dbCORTHot, probright_dbControlHot)) / mean(c(probright_drCORTCold, probright_drControlCold, probright_drCORTHot, probright_drControlHot))) - 1) * 100, 1)`% increase in the probability of choosing the right ramp on trial 1 compared to lizards assigned to the red ramps (mean P~Blue~ = `r format_dec(mean(c(probright_dbCORTCold, probright_dbControlCold, probright_dbCORTHot, probright_dbControlHot)), 3)`, 95% CI P~Blue~ = [`r format_dec(quantile(c(probright_dbCORTCold, probright_dbControlCold, probright_dbCORTHot, probright_dbControlHot), 0.025), 3)` , `r format_dec(quantile(c(probright_dbCORTCold, probright_dbControlCold, probright_dbCORTHot, probright_dbControlHot), 0.975), 3)`]; mean P~Red~ = `r format_dec(mean(c(probright_drCORTCold, probright_drControlCold, probright_drCORTHot, probright_drControlHot)), 3)`, 95% CI P~Red~ = [`r format_dec(quantile(c(probright_drCORTCold, probright_drControlCold, probright_drCORTHot, probright_drControlHot), 0.025), 3)` , `r format_dec(quantile(c(probright_drCORTCold, probright_drControlCold, probright_drCORTHot, probright_drControlHot), 0.975), 3)`]), suggesting a bias towards blue. A blue bias was also supported by initial choices differing significantly from chance, and this was consistent across all treatments (i.e., P > 0.33; Table 2 in _Supplementary Material_). However, this difference in the probability of correctly choosing between red and blue ramps was not significant, likely due to small sample sizes (Contrast between probabilities: P~Blue~ - P~Red~ = `r format_dec(mean(c(probright_dbCORTCold, probright_dbControlCold, probright_dbCORTHot, probright_dbControlHot)) - mean(c(probright_drCORTCold, probright_drControlCold, probright_drCORTHot, probright_drControlHot)), 3)`, p~mcmc~ `r format_p(pmcmc(c(probright_dbCORTCold, probright_dbControlCold, probright_dbCORTHot, probright_dbControlHot) - c(probright_drCORTCold, probright_drControlCold, probright_drCORTHot, probright_drControlHot)), 3)`, see @fig-deli).

>>>>

## Abstract
Animals must acquire new information through learning to adjust their behavior adaptively. However, learning ability can be constrained by conditions experienced during early development, when the brain is especially susceptible to environmental conditions. For example, temperature can result in phenotypically plastic adjustments to growth, metabolism, and learning in ectotherms. In vertebrates, thermal conditions can increase the production of glucocorticoid (GCs) - ‘stress’ hormones. Maternal GCs can be transmitted to offspring during development, potentially impacting their learning abilities. GCs and thermal environments are, therefore, predicted to have interactive effects on the development of learning in ectotherms. Here, we investigated the combined effects of prenatal corticosterone (CORT) - the main GC in reptiles - and incubation temperature on associative learning using two species of lizards, _Lampropholis delicata_ and _L. guichenoti_. We manipulated CORT levels and temperature in a 2x2 factorial design, and then subjected juveniles to a color-associative learning task. We predicted that elevated CORT and low temperatures would impair associative learning. However, both species showed similar learning rates independently of treatment. Our results suggest that these two species may have evolved mechanisms to maintain learning performance despite prenatal challenges. We also found that color affected decision-making in both species. Overall, we observed a non-learned preference towards blue, underscoring the need to carefully select the color used in cognitive tests involving visual stimuli.

## Introduction
Cognition is a set of processes by which animals gather, preserve, and use information from their environment through perception, learning, memory, and decision-making [@shettleworth]. These processes are fundamental to foraging, mate selection, antipredatory strategies, and social behaviors, all of which are crucial for survival and reproduction [@dukas_evolutionary_2004]. Learning - acquiring and consolidating new information [@dukas_evolutionary_2004] - is essential for coping with environmental changes by enabling individuals to create new associations between events [@dukas_evolutionary_2004; @leal_behavioural_2012; @buchanan_condition_2013]. However, the capacity to acquire information varies among individuals, shaped by factors such as age, sex, and the developmental environment [@szuran_water_1994; @lemaire_prenatal_2000; @zhu_prenatal_2004; @amiel_hotter_2012; @amiel_egg_2014; @carazo_sex_2014; @noble_age-dependent_2014]. Interindividual differences in learning can lead to unequal responses to environmental conditions, potentially affecting population dynamics [@ward2016ecological; @welklin2024spatial]. For example, fast learners may better exploit novel resources or avoid new dangerous stimuli, while those with lower learning capabilities might struggle to adapt to environmental changes [@ward2016ecological]. Learning rate could affect their survival and reproductive output, ultimately influencing population growth rates and stability [@ward2016ecological; @welklin2024spatial]. Therefore, understanding the factors that shape learning is crucial to predict how populations will respond to novel conditions.

@@

Because the brain is highly susceptible to environmental inputs during early stages of life, developmental conditions are especially relevant in shaping cognitive abilities  [@zhu_prenatal_2004]. Impacts on brain development can have long-lasting effects on cognition, potentially influencing an individual's capacity to learn and adapt to new environments [@lemaire_prenatal_2000; @zhu_prenatal_2004; @amiel_hotter_2012; @abayarathna_effects_2020]. For example, the state of the mother can influence offspring phenotype beyond genetic transmission through nest-site selection, provisioning, or the transmission of hormones or nutrients [i.e., 'maternal effects' @moore_mothers_2019]; and the developing brain is also susceptible to these effects.

Glucocorticoids (GCs) are a class of steroid hormones that are particularly relevant in phenotypic plasticity. In vertebrates, GCs regulate metabolism and maintain homeostasis in response to disturbances [i.e., ‘the stress response’ @sapolsky_how_2000; @picard_mitochondrial_2014]. Under stressful situations, animals react by initiating adaptive physiological and behavioral adjustments mediated by GCs. These GCs can be transmitted directly from parents to their offspring with various effects on offspring phenotype [reviewed by @Crino_2023]. Elevation in GCs during early stages of development typically results in altered neurogenesis, brain structure, and metabolic activity that, in most cases, are related to decreased learning abilities [@lemaire_prenatal_2000; @zhu_prenatal_2004; @du_dynamic_2009; @eaton2015mild; @farrell_developmental_2015-learn; @farrell_developmental_2016]. For instance, prenatal stress in rats (_Rattus norvegicus_) suppresses neurogenesis in the dentate gyrus, which is associated with impairments in hippocampal-related spatial tasks [@lemaire_prenatal_2000]. Factors such as sex or the nature of the learning task can also affect the direction and magnitude of the effects of prenatal exposure to elevated GCs [@szuran_water_1994; @crino_corticosterone_2014-learn; @farrell_developmental_2015-learn; @bebus_associative_2016; @farrell_developmental_2016]. Because the effects of prenatal GCs on learning can be context-dependent, it is crucial to conduct studies across diverse taxa and experimental conditions to understand these effects fully.

In addition to the environments experienced by parents, offspring also experience potentially stressful environmental conditions that can interact with or amplify parental effects. In ectotherms, the early thermal environment is a mechanism of phenotypic plasticity, influencing a broad spectrum of traits, including growth, metabolism, or cognition [@amiel_hotter_2012; @amiel_egg_2014; @dayananda_incubation_2017; @noble_developmental_2018; @abayarathna_effects_2020]. For instance, Port Jackson sharks (_Heterodontus portusjacksoni_) incubated at warmer temperatures took fewer days to master a numerical learning task than those incubated at cooler [@vila_pouca_quantity_2019]. In skinks, high incubation temperatures are generally associated with faster learning rates [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014]; but velvet geckos (_Amalosia lesueurii_) incubated at temperatures beyond their natural thermal range are worse learners than those incubated at colder temperatures [@dayananda_incubation_2017; @abayarathna_effects_2020]. Changes in neural structure and function likely mediate the effects of incubation temperature in reptiles, as high temperatures increase neural density and metabolic activity in the brain [@coomber_independent_1997; @sakata_neural_2000; @amiel_effects_2017; @beltran_are_2021].

GCs can play a pivotal role in determining vertebrate responses to elevated temperatures [@Crino_2023], potentially fostering natural interactions between temperature and GCs. Additionally, GCs and temperatures can act upon similar physiological mechanisms [@coomber_independent_1997; @lemaire_prenatal_2000; @sakata_neural_2000; @zhu_prenatal_2004; @du_dynamic_2009; @amiel_effects_2017; @beltran_are_2021]. Therefore, the effects of GCs and temperature may be interdependent. However, the interactive effects of GCs and temperature on learning abilities remain unexplored, yet understanding of how temperature and GC’s interact is particularly relevant in global climate change [@vinogradov2024inbreeding].

Here, we investigated the combined effects of prenatal corticosterone (CORT) - the main GC in reptiles - and incubation temperature on cognition in two species of skinks, the delicate skink (_Lampropholis delicata_) and the common garden skink (_L. guichenoti_). We manipulated CORT levels in the eggs and then incubated them at two different temperatures in a 2X2 factorial design. Post-incubation, juveniles were subjected to a color-associative task to assess their learning abilities. We hypothesized that prenatal CORT levels and thermal environment would impact the learning of an association task. We predicted that individuals exposed to high levels of CORT or low temperatures would learn slower than control individuals or those incubated at higher temperatures [@zhu_prenatal_2004; @amiel_hotter_2012; @eaton2015mild; @amiel_effects_2017]. Additionally, we predicted that incubation at high temperatures would mitigate the impact of CORT on learning performance, while cold incubation temperatures were expected to enhance the effects of CORT. The interactive effects of CORT and temperature may occur for two reasons that are not mutually exclusive: first, an increase in temperature leads to an overall rise in neural density [@amiel_effects_2017], thereby counteracting the impact of CORT [see @lemaire_prenatal_2000; @zhu_prenatal_2004; @eaton2015mild]; and second, the elevated metabolic rate associated with higher temperatures could accelerate CORT metabolism, resulting in embryos being exposed to CORT for a shorter time.

@@

*Eggs collection and incubation* -- Between mid-October 2022 to the end of February 2023, we provided females with a place to lay eggs by means of small boxes (12.5 L x 8.3 W x 5 H cm) containing moist vermiculite. These boxes were placed in one side of the communal enclosures (see above). We checked for eggs in the boxes three days a week (Monday, Wednesday, and Friday). After collection, we measured the length and width of eggs with a digital caliper to the nearest 0.1 mm, and mass using a digital scale (OHAUS, Model spx123; ± 0.001 g error). Then, we assigned individual IDs to the clutch and each egg. Eggs were treated with a CORT or control treatment (see CORT and Temperature Manipulation below) and were placed in individual cups (80 mL) with moist vermiculite (12 parts water to 4 parts vermiculite). The cups were covered with cling wrap to retain moisture and left in LATWIT 2X5D-R1160 incubators at two different temperatures (see CORT and Temperature manipulation below) until hatching. Eggs were then checked three times a week for hatchlings.

*Hatchlings* -- Immediately after hatching, we measured lizards' snout-to-vent length (SVL) and tail length (TL) with a ruler to the nearest mm, and mass using a digital scale (OHAUS, Model spx123; ± 0.001 g error). Hatchlings were then housed individually in small enclosures (18.7L x 13.2W x 6.3H cm) provided with non-stick matting, shelter, and a water dish. All care otherwise follows similar protocols to adults (see above).

Two weeks prior to the training phase (see details below), 1.5 to 5 months old lizards (see details in Statistical Analyses and Supplementary Material) were moved to the experimental arena for acclimatization. The arenas were individual medium-size (41 L x 29.7 W x 22 H cm) plastic containers with a shelter (9 L x 6 W x 1.5 H cm) on one of the sides and a water dish on the other. These arenas were placed in two rooms in 7 different racks that were monitored by 7 different CCTV systems (device model DVR-HP210475), allowing us to record their behavior during the experiment (see details below). Although the conditions in the experimental rooms were identical to the colony room, the number of lizards per species and treatment in each rack was counterbalanced to control for any potential effect of the room or the position of the lizard in the rack. During acclimatization and throughout the experiment, lizards were fed with only one cricket per day dusted with calcium and multivitamins, and water was supplied *ad libitum*. We provided a temperature gradient by means of a heat cord and heat lamps in a 12 h light: 12 h dark cycle.

#### CORT and Temperature Manipulation  
To empirically test the effect of early environment, we manipulated CORT concentration in eggs and incubated them under one of two temperature regimes (‘Cold’ - 23 ± 3 ºC or ‘Hot’ - 28 ± 3 ºC) in a 2x2 factorial design (@fig-Methods A). We topically dosed eggs with a CORT solution (10 pg/mL) or a control treatment (100% ethanol). Corticosterone treatments were made by dissolving crystalline corticosterone (Sigma, Cat. No. C2505) in 100% ethanol. To dose eggs, we applied 5µl of solutions to eggshells using a micropipette. We selected these doses based on previous studies publishing yolk CORT concentrations in other oviparous reptiles [@lovern2008effects; @hanover2019corticosterone], while also validating that it fell within the range of CORT concentrations in eggs within our population [@crino2024eggs]. CORT treatment increased mean yolk CORT levels ~3.7x higher than control eggs [@crino2024eggs]. After the treatment, the eggs were incubated in one of the two previously mentioned temperature regimes (‘Cold’ or ‘Hot’) until hatching. The number of eggs per clutch assigned to each hormone and temperature treatment was counterbalanced in both species.

#### Learning  
To estimate learning skills, we tested each lizard's ability to locate a food reward in a color-associative learning task (@fig-Methods C, D). First, we performed a training phase where lizards had to learn to eat from white 3D-printed PLA ramps (9 L x 4 W x 5 H cm) identical in size and shape to the ones used during the associative task (@fig-Methods B). We divided the training phase into three stages: the first stage where lizards had to eat a small, frozen cricket (*A. domestica*) from an opaque petri dish (3D x 1.6H cm) placed in the middle of their enclosure (@fig-Methods C, Stage 1); the second stage where the petri dish with the cricket was placed on top of the white 3D printed ramps (@fig-Methods C, Stage 2); and the third where the cricket was left inside a well (3D x 1.75H cm) on the top of the ramp (@fig-Methods C, Stage 3). Every trial began when we left the feeding block (petri dish, ramp, or both) inside the enclosure and finished one hour later when we took it away. At the end of each trial, we recorded whether the cricket had been consumed. A trial was considered successful if the lizard could locate and consume the reward, while completion of each stage required the lizards to eat the crickets in at least 5 out of 6 trials to ensure lizards were feeding consistently. This phase lasted 38 days until all the lizards learned to eat from the ramp; only in one case did we decide not to use the lizard because its behavior inconsistent throughout  the course of the training phase.

@@

```{r, fig-Methods}
#| label: fig-Methods
#| fig.cap: "Experimental design of the study. Panel A shows the early environment manipulation. In panel B, the measurements of the 3D-printed ramps that were employed in the habituation and learning tasks. Panel C shows the habituation process in three different stages. In panel D, the associative task is done with the three different 3D-printed ramps. White lids in D show the ramps where the food reward was not attainable."

@@

We started with 96 lizards, 48 per species and 12 per treatment per species. However, due to mortality (n = 11), failing to pass the training stage (n = 1), or lack of motivation during the learning tasks (n = 3), we had a final sample size of 81 lizards. Final sample sizes per treatment and species are listed in @fig-deli and @fig-guich (figures for both species with the raw data are included in the _Supplementary Material_). Mean slopes (denoted as β throughout) per treatment for both species are provided in Table 1 in the _Supplementary Material_.    
```{r, models}
#| label: models
# Fitting the model and extraction of posteriors for both species using fit_m function (see func.r in R folder). The result everytime the function is used is a df with the posteriors of the model. The functions saves the model automatically in output/models; and when the parameter refit = FALSE then the posteriors are extracted from the model previously written instead of fitting the model again each time.
source(here("R", "func.R"))
# Model formula: FC_associative ~ Trial*cort*temp + (1 + Trial|lizard_id) + (1| clutch)
## A.1) L. delicata
deli_red <- fit_m("deli", "red", refit = FALSE)
write.csv(deli_red, file= "./output/Checking/deli_red.csv")
deli_blue <- fit_m("deli", "blue", refit = FALSE)
write.csv(deli_blue, file= "./output/Checking/deli_blue.csv")
## A.2) L. guichenoti
guich_red <- fit_m("guich", "red", refit = FALSE)
write.csv(guich_red, file= "./output/Checking/guich_red.csv")
guich_blue <- fit_m("guich", "blue", refit = FALSE)
write.csv(guich_blue, file= "./output/Checking/guich_blue.csv")
```

@@

_Lampropholis delicata_: individuals given the blue ramp as the correct choice had an `r format_dec(((mean(c(probright_dbCORTCold, probright_dbControlCold, probright_dbCORTHot, probright_dbControlHot)) / mean(c(probright_drCORTCold, probright_drControlCold, probright_drCORTHot, probright_drControlHot))) - 1) * 100, 1)`% increase in the probability of choosing the right ramp on trial 1 compared to lizards assigned to the red ramps (mean P~Blue~ = `r format_dec(mean(c(probright_dbCORTCold, probright_dbControlCold, probright_dbCORTHot, probright_dbControlHot)), 3)`, 95% CI P~Blue~ = [`r format_dec(quantile(c(probright_dbCORTCold, probright_dbControlCold, probright_dbCORTHot, probright_dbControlHot), 0.025), 3)` , `r format_dec(quantile(c(probright_dbCORTCold, probright_dbControlCold, probright_dbCORTHot, probright_dbControlHot), 0.975), 3)`]; mean P~Red~ = `r format_dec(mean(c(probright_drCORTCold, probright_drControlCold, probright_drCORTHot, probright_drControlHot)), 3)`, 95% CI P~Red~ = [`r format_dec(quantile(c(probright_drCORTCold, probright_drControlCold, probright_drCORTHot, probright_drControlHot), 0.025), 3)` , `r format_dec(quantile(c(probright_drCORTCold, probright_drControlCold, probright_drCORTHot, probright_drControlHot), 0.975), 3)`]), suggesting a bias towards blue. A blue bias was also supported by initial choices differing significantly from chance, and this was consistent across all treatments (i.e., P > 0.33; Table 2 in _Supplementary Material_). However, this difference in the probability of correctly choosing between red and blue ramps was not significant, likely due to small sample sizes (Contrast between probabilities: P~Blue~ - P~Red~ = `r format_dec(mean(c(probright_dbCORTCold, probright_dbControlCold, probright_dbCORTHot, probright_dbControlHot)) - mean(c(probright_drCORTCold, probright_drControlCold, probright_drCORTHot, probright_drControlHot)), 3)`, p~mcmc~ `r format_p(pmcmc(c(probright_dbCORTCold, probright_dbControlCold, probright_dbCORTHot, probright_dbControlHot) - c(probright_drCORTCold, probright_drControlCold, probright_drCORTHot, probright_drControlHot)), 3)`, see @fig-deli).

Decision-making by _L. delicata_ was not impacted by the CORT (P~Control~ - P~CORT~ = `r format_dec(mean(c(probright_drControlCold, probright_drControlHot)) - mean(c(probright_drCORTCold, probright_drCORTHot)), 3)`, p~mcmc~ `r format_p(pmcmc(c(probright_drControlCold, probright_drControlHot) - c(probright_drCORTCold, probright_drCORTHot)), 3)`), incubation temperature (P~Hot~ - P~Cold~ = `r format_dec(mean(c(probright_drControlHot, probright_drCORTHot)) - mean(c(probright_drControlCold, probright_drCORTCold)), 3)`, p~mcmc~ `r format_p(pmcmc(c(probright_drControlHot, probright_drCORTHot) - c(probright_drControlCold, probright_drCORTCold)), 3)`) or their interaction ([(P~Control-Hot~ - P~CORT-Hot~) - (P~Control-Cold~ - P~CORT-Cold~)] = `r format_dec(mean(probright_drControlHot - probright_drCORTHot) - mean(probright_drControlCold - probright_drCORTCold), 3)`, p~mcmc~ `r format_p(pmcmc(probright_drControlHot - probright_drCORTHot - (probright_drControlCold - probright_drCORTCold)), 3)`) when red ramps were the correct choice (see @fig-deli A). When blue ramps were the correct choice, there was no significant effect of hormone (P~Control~ - P~CORT~ = `r format_dec(mean(c(probright_dbControlCold, probright_dbControlHot)) - mean(c(probright_dbCORTCold, probright_dbCORTHot)), 3)`, p~mcmc~ `r format_p(pmcmc(c(probright_dbControlCold, probright_dbControlHot) - c(probright_dbCORTCold, probright_dbCORTHot)), 3)`), temperature (P~Hot~ - P~Cold~ = `r format_dec(mean(c(probright_dbControlHot, probright_dbCORTHot)) - mean(c(probright_dbControlCold, probright_dbCORTCold)), 3)`, p~mcmc~ `r format_p(pmcmc(c(probright_dbControlHot, probright_dbCORTHot) - c(probright_dbControlCold, probright_dbCORTCold)), 3)`) or their interaction ([(P~Control-Hot~ - P~CORT-Hot~) - (P~Control-Cold~ - P~CORT-Cold~)] = `r format_dec(mean(probright_dbControlHot - probright_dbCORTHot) - mean(probright_dbControlCold - probright_dbCORTCold), 3)`, p~mcmc~ `r format_p(pmcmc(probright_dbControlHot - probright_dbCORTHot - (probright_dbControlCold - probright_dbCORTCold)), 3)`) either (see @fig-deli D).
